---
title: "Getting Started with MCPR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with MCPR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Getting Started with MCPR: Your First Session

MCPR (Model Context Protocol for R) gives your AI assistant a persistent R workspace. By default, that workspace is the private R process running `MCPR::mcpr_server()`. When you want the agent to work inside an existing interactive R console, you can make that console attachable with `mcpr_session_start()` and switch the active session with `manage_r_sessions()`.

## Installation and Setup

```{r eval=FALSE}
# Install MCPR
if (!require("remotes")) install.packages("remotes")
remotes::install_github("phisanti/MCPR")

# Configure your AI agent (Claude, Gemini, or Copilot)
library(MCPR)
install_mcpr(agent = "claude")
```

## Your First MCPR Experience

Let's walk through a basic session to see how MCPR works in practice.

### Step 1: Use the Private Session

```{r eval=FALSE}
# Your MCP client launches this through its configuration
MCPR::mcpr_server()
```

The MCP server process is also the agent's private R session. Variables created by ordinary tools persist there until the server process exits.

### Step 2: Check the Workspace

Now you can ask your AI assistant:

> **You:** "What variables are in the R workspace?"

Your AI will use MCPR tools to respond:

```r
# Check current variables
ls()
# Shows objects in the private session
```

### Step 3: Create Some Variables

> **You:** "Can you create some random variables?"

Your AI assistant executes:

```r
# Create multiple variables efficiently
random_numbers <- runif(15, 1, 100)
random_matrix <- matrix(rnorm(12), nrow = 3, ncol = 4)
random_names <- sample(c("Alice", "Bob", "Charlie", "Diana", "Eve"), 8, replace = TRUE)
random_df <- data.frame(
  id = 1:6,
  value = rnorm(6, mean = 50, sd = 10),
  category = sample(letters[1:3], 6, replace = TRUE)
)

# Verify creation
ls()
```

**Key insight:** All these variables now persist in the private session and remain available for future operations.

### Step 4: Optionally Attach a Live R Session

> **You:** "I want you to work inside my existing R console"

In that R console, run:
```{r eval=FALSE}
library(MCPR)
mcpr_session_start()
```

Then ask your AI:

> **You:** "List attachable sessions and join mine"

Your AI will:

```r
manage_r_sessions("list")
manage_r_sessions("join", session = 1)
```

Ordinary tool calls still omit `session`; after joining, MCPR sends them to the active attached session. Use `manage_r_sessions("detach")` to return to the private session.

### Step 5: Session Independence

**Key discovery:** The private session, human sessions, and MCPR-owned secondary sessions maintain independent workspaces. Attaching changes the active execution target; it does not merge workspaces.

## Core MCPR Tools

### Session Management
```{r eval=FALSE}
# List all active sessions
manage_r_sessions("list")

# Join a specific session
manage_r_sessions("join", session = 2)

# Return to the private session
manage_r_sessions("detach")
```

### Code Execution
```{r eval=FALSE}
# Execute R code in the active session
execute_r_code("
  summary(random_df)
  plot(random_numbers)
")
```

### Workspace Inspection
```{r eval=FALSE}
# View current objects
view("session")

# View command history
view("terminal")

# View file system
view("workspace")
```

### Plot Creation
```{r eval=FALSE}
# Show a plot to the user
show_plot("
  hist(random_numbers, main = 'Random Numbers Distribution')
")
```

## Conclusion

This little demo aims to show the capabilities of MCPR. It provides a persistent R environment that bridges the gap between traditional scripting and interactive AI agent collaboration:

- **Stateful Programming**: Your AI agent maintains context in a persistent private workspace
- **Parallel Workflows**: Private, human, and secondary sessions let you compartmentalize different projects
- **Controlled Attachment**: Move between sessions with `manage_r_sessions` while ordinary tools stay simple
- **Interactive Analysis**: Discuss data, iterate on visualizations, and explore findings conversationally
- **Persistent Collaboration**: Build complex analyses incrementally with your AI agent as a programming partner

MCPR transforms R from isolated script execution into a collaborative workspace where your AI agent becomes an extension of your analytical thinking—whether you're exploring datasets, debugging complex models, or building reproducible research workflows.

## Troubleshooting

**"No attachable sessions found"**: If you want to attach to a live R console, start one with `mcpr_session_start()`

**"Connection failed"**: Verify MCPR installation with `install_mcpr()`

**"Variables not found"**: Check the active session with `manage_r_sessions("list")`, or detach back to private/local execution with `manage_r_sessions("detach")`

## Next Steps

- **[AI Agent Integration](ai-agent-integration.html)**: Configure your specific AI agent
- **[Tool Development](tool-development.html)**: Create custom tools for your workflow
- **[Architecture Overview](architecture.html)**: Understand how MCPR works internally

You're now ready to use MCPR for persistent, collaborative R programming with your AI assistant!
