---
title: "MCP Setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MCP Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# AI Agent Integration Guide

This guide covers how to integrate MCPR with different AI coding assistants.

## Supported Agents

MCPR currently supports:
- Claude Desktop
- Gemini (via MCP)
- GitHub Copilot (via MCP)

## Claude Desktop Setup

### Automatic Configuration
```{r eval=FALSE}
library(MCPR)
install_mcpr(agent = "claude")
```

### Manual Configuration
Edit your Claude Desktop config file:

**macOS**: `~/Library/Application Support/Claude/claude_desktop_config.json`
**Windows**: `%APPDATA%\Claude\claude_desktop_config.json`  
**Linux**: `~/.config/claude/claude_desktop_config.json`

Add the MCP server configuration:

```json
{
  "mcpServers": {
    "mcpr": {
      "command": "R",
      "args": ["--quiet", "--slave", "-e", "MCPR::mcpr_server()"]
    }
  }
}
```

## Using MCPR Tools

Once configured, your AI agent will have access to these tools:

### Session Management
- `manage_r_sessions('list')` - Show the private session, active session, and attachable sessions
- `manage_r_sessions('join', session=ID)` - Attach to a human session started with `mcpr_session_start()`
- `manage_r_sessions('start')` - Start and attach an MCPR-owned secondary session
- `manage_r_sessions('detach')` - Return ordinary tools to the private/local session

### Code Execution  
- `execute_r_code(code)` - Run R code in the active session. Do not pass a session argument.

### Visualization
- `show_plot(expr, target, ...)` - Create and display plots (target='user' or 'agent')

### Inspection
- `view('session')` - View workspace objects
- `view('terminal')` - View command history  
- `view('workspace')` - View file system context

## Workflow Examples

### Data Analysis Workflow
1. The MCP client starts `MCPR::mcpr_server()`.
2. Load data in the private session: `execute_r_code("data <- read.csv('data.csv')")`
3. Explore: `view('session')` to see loaded objects
4. Visualize: `show_plot("plot(data$x, data$y)")`
5. Optionally attach to a live R console with `manage_r_sessions('join', session=ID)`
6. Detach back to private/local execution with `manage_r_sessions('detach')`

### Multi-Session Development
- Private session: agent-owned package work
- Human session: live interactive console made attachable with `mcpr_session_start()`
- Secondary session: MCPR-owned process created by `manage_r_sessions('start')`

## Troubleshooting

**Agent can't see tools**: Restart the agent after configuration.

**Permission errors**: Check file paths and R library permissions.

**Connection timeouts**: Increase timeout in MCP configuration.

## Advanced Configuration

### Custom Tool Paths
```json
{
  "mcpServers": {
    "mcpr": {
      "command": "R",
      "args": ["--quiet", "--slave", "-e", "MCPR::mcpr_server()"],
      "env": {
        "MCPTOOLS_TOOLS_DIR": "/path/to/custom/tools",
        "MCPTOOLS_LOG_FILE": "/path/to/logs/mcpr.log"
      }
    }
  }
}
```

### Environment Variables
- `MCPTOOLS_LOG_FILE`: Path for debug logs
- `MCPTOOLS_TOOLS_DIR`: Custom tools directory
- `R_LIBS_USER`: R library path

## Common Issues

### Working with MCPR in renv Projects

When using MCPR with AI agents (like Claude Desktop or Codex) in renv-enabled projects, you may encounter connection issues where the agent cannot find or load the MCPR package.

**Problem**: The MCP server fails to start or behaves unexpectedly because `renv` isolates the project's R library. This can cause the AI agent's private MCPR server process to use a different instance or version of `MCPR` than an interactive R console you later attach.

**Solution**: Configure renv to use your global MCPR installation by creating a scaffold file before initializing renv:

1. **Before running `renv::init()`**, create a file called `scaffold.R` in your project root:

```r
# scaffold.R
library(MCPR)
```

2. Initialize renv normally:

```r
renv::init()
```

This approach ensures that renv detects MCPR as a dependency and links to your global installation rather than trying to install a separate copy. This is similar to the [recommended approach for VSCode-R with renv](https://github.com/REditorSupport/vscode-R/wiki/Working-with-renv-enabled-projects).

**For existing renv projects**:

If you already have an renv project without MCPR, you can add it:

```r
# Install MCPR into the renv library
renv::install("path/to/MCPR", repos = NULL, type = "source")

# Or use the global installation
renv::use_package("MCPR", type = "Imports", dev = TRUE)
renv::snapshot()
```

### GitHub Copilot / Codex Configuration Issues

Some newer versions of AI coding assistants require explicit specification of the MCP server connection type.

**Problem**: MCP server fails to connect with error messages like "handshaking with MCP server failed" or "connection closed: initialize response".

**Solution**: Ensure your configuration file explicitly declares `"type": "stdio"`.

**Correct configuration** for `~/.vscode/mcp.json` or equivalent:

```json
{
    "servers": {
        "MCPR": {
            "type": "stdio",
            "command": "R",
            "args": [
                "--quiet",
                "--slave",
                "-e",
                "MCPR::mcpr_server()"
            ]
        }
    },
    "inputs": []
}
```

**Key points**:
- The `"type": "stdio"` field is **required** for newer MCP client implementations
- Without this field, the client may fail to establish the communication channel
- The `"inputs": []` field may also be required by some clients

### Protocol Version Mismatches

Different AI coding agents support different MCP protocol versions. MCPR implements protocol version negotiation to handle this automatically.

**Supported versions**:
- `2024-11-05` (baseline)
- `2025-03-26`
- `2025-06-18`
- `2025-11-25` (latest)

If you encounter version-related errors, check your agent's logs for protocol version information. MCPR should automatically negotiate the highest mutually supported version.
