---
title: "Composing Tool Surfaces with MCPR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Composing Tool Surfaces with MCPR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Building Domain-Specific MCP Servers on MCPR

MCPR is designed to be a substrate. It handles transport, private-session execution, optional session attachment, and tool schemas so that domain packages can focus on what they know: their own tools. This vignette shows how `ToolRegistry` supports that through multi-directory discovery and filtering.

## The Problem

By default, `mcprServer$new()` discovers all tools shipped with MCPR:

```{r}
library(MCPR)

registry <- ToolRegistry$new(tools_dir = system.file(package = "MCPR"))
registry$search_tools()
registry$get_tool_summary()
#>                  name                                        description parameters source_dir
#> 1     execute_r_code     Execute R code in the active R session and ...          2  /path/MCPR
#> 2     inspect_object     Inspect an R object in the active session ...          3  /path/MCPR
#> 3 manage_r_sessions      Manage R sessions: list, join, or get info...          2  /path/MCPR
#> 4          show_plot     Generate and display R plots as base64 PNG...          4  /path/MCPR
#> 5               view     View the current state of the R session, t...          2  /path/MCPR
```

That's the right default for a general-purpose R assistant. But a domain package like `mypackage` wants a narrower surface: its own tools, maybe optional attachment controls, and definitely *not* `execute_r_code` — which would let an LLM run arbitrary code in a production pipeline.

Two mechanisms solve this: **multi-directory discovery** and **filtering**.

## Case 1: MCPR Tools Only (Default)

Nothing changes. This is how MCPR works today:

```{r}
server <- mcprServer$new()
```

Under the hood this creates a `ToolRegistry` pointed at MCPR's installed `inst/` directory, calls `search_tools()`, and loads every `tool-*.R` file it finds.

## Case 2: MCPR Tools + Your Package's Tools

Pass a character vector to `tools_dir`:

```{r}
mcpr_dir <- system.file(package = "MCPR")
my_dir   <- system.file("tools", package = "mypackage")

registry <- ToolRegistry$new(
  tools_dir = c(mcpr_dir, my_dir)
)
registry$search_tools()
registry$get_tool_summary()
#>                  name                                        description parameters   source_dir
#> 1     execute_r_code     Execute R code in the active R session and ...          2   /path/MCPR
#> 2     inspect_object     ...                                                    3   /path/MCPR
#> ...
#> 7                add     Add two numbers and return the result.                  2  /path/mypackage
#> 8           subtract     Subtract b from a and return the result.                2  /path/mypackage

server <- mcprServer$new(registry = registry)
```

Each directory is scanned independently. If a directory doesn't exist it's skipped with a warning, not an error — so your package works gracefully whether a dependency is installed or not.

**Duplicate tool names produce a warning.** If both directories define a tool called `view`, `search_tools()` warns with provenance details showing which directories contributed the conflict. You should resolve it by filtering one out — silent shadowing causes debugging nightmares in composed registries.

## Case 3: Only Your Package's Tools

Point the registry at your package alone:

```{r}
registry <- ToolRegistry$new(
  tools_dir = system.file("tools", package = "mypackage")
)
registry$search_tools()

server <- mcprServer$new(registry = registry)
```

No MCPR built-in tools are loaded. Your server exposes exactly what your package defines and runs local-only in the private MCP server process. There is no hidden session discovery or attachment behavior.

## Case 4: Cherry-Pick from MCPR + All of Your Package

This is the most common real-world pattern. You want MCPR's optional session-management control plane, but not `execute_r_code` or other open-ended tools.

Two approaches, same result:

**Approach A — exclude what you don't want:**

```{r}
registry <- ToolRegistry$new(
  tools_dir = c(system.file(package = "MCPR"), system.file("tools", package = "mypackage"))
)
registry$search_tools()
registry$filter(exclude = c("execute_r_code", "show_plot", "inspect_object"))

registry$get_tool_summary()
#>                 name         ...   source_dir
#> 1 manage_r_sessions         ...   /path/MCPR
#> 2              view         ...   /path/MCPR
#> 3               add         ...   /path/mypackage
#> 4          subtract         ...   /path/mypackage

server <- mcprServer$new(registry = registry)
```

**Approach B — include what you want:**

```{r}
registry <- ToolRegistry$new(
  tools_dir = c(system.file(package = "MCPR"), system.file("tools", package = "mypackage"))
)
registry$search_tools()
registry$filter(include = c("manage_r_sessions", "add", "subtract"))
```

Use `exclude` when you want "everything except these". Use `include` when you want "only these, nothing else".

Filters persist across `search_tools()` calls — if you re-scan (e.g., `search_tools(force_refresh = TRUE)`), the same filter is reapplied automatically. Call `filter()` with no arguments to clear all filters.

## Case 5: Multiple Tool Subdirectories from One Package

Some packages organize tools by domain:

```
mypackage/
  inst/
    tools-basic/
      tool-add.R
      tool-subtract.R
    tools-advanced/
      tool-multiply.R
      tool-divide.R
```

Point the registry at all of them:

```{r}
pkg_base <- system.file(package = "mypackage")

registry <- ToolRegistry$new(
  tools_dir = c(
    file.path(pkg_base, "tools-basic"),
    file.path(pkg_base, "tools-advanced")
  )
)
registry$search_tools()
registry$get_tool_summary()
#>        name                              ...     source_dir
#> 1       add   Add two numbers.                /path/tools-basic
#> 2  subtract   Subtract b from a.              /path/tools-basic
#> 3  multiply   Multiply two numbers.           /path/tools-advanced
#> 4    divide   Divide a by b.                  /path/tools-advanced
```

Or load a subset of modules:

```{r}
# Basic calculator only — no advanced operations
registry <- ToolRegistry$new(
  tools_dir = file.path(pkg_base, "tools-basic")
)
```

## Chaining: configure(), filter(), search_tools()

`configure()`, `filter()`, and `set_verbose()` return `self` (invisibly) for chaining. `search_tools()` returns the tool list directly. The intended usage pattern is:

```{r}
# Declarative: set up, then scan
registry <- ToolRegistry$new(tools_dir = system.file(package = "MCPR"))
registry$search_tools()
registry$filter(exclude = "execute_r_code")

# Or reconfigure an existing registry
registry$configure(
  tools_dir = c(system.file(package = "MCPR"), system.file("tools", package = "mypackage"))
)
registry$search_tools()
registry$filter(include = c("manage_r_sessions", "add"))
```

`configure()` clears the cache and resets the directory list. `filter()` applies immediately to already-discovered tools *and* persists for future scans. `search_tools()` respects both.

## Provenance: Where Did This Tool Come From?

Every discovered tool is stamped with `source_dir` and `source_file` in its annotations. This shows up in `get_tool_summary()` and is accessible programmatically:

```{r}
tool <- registry$get_tool("add")
tool$annotations$source_dir
#> [1] "/Library/R/4.3/library/mypackage/tools-basic"
tool$annotations$source_file
#> [1] "tool-add.R"
```

Enable verbose mode to see provenance during discovery:

```{r}
registry$set_verbose(TRUE)
registry$search_tools(force_refresh = TRUE)
#> ℹ Searching 4 files across 2 directories...
#> ℹ Loaded 1 tool from tool-add.R
#> ℹ Loaded 1 tool from tool-subtract.R
#> ...
#> ℹ Found 4 tools, 2 active after filtering.
```

## Writing Tools for Your Package

MCPR discovers tools through roxygen2 tags. A tool file in your package looks exactly like MCPR's own:

```{r, code = readLines(system.file("tool-view.R", package = "MCPR"))[1:15], eval = FALSE}
```

The key elements:

1. **`@keywords mcpr_tool`** — this is how `ToolRegistry` identifies tool functions
2. **`@param`** tags with type annotations — converted to JSON Schema for MCP clients
3. **`@description`** — becomes the tool description that LLMs see

### Required vs Optional Parameters

MCPR now follows two different patterns depending on how the tool is defined:

- **Roxygen auto-discovery (`ToolRegistry`)** infers the JSON Schema `required`
  array from the function signature. Parameters **without** a default value are
  emitted as required; parameters **with** a default value are emitted as
  optional. This matches the signature-driven pattern used by the official MCP
  Python SDK.
- **Explicit `tool()` definitions** keep the declared `mcpr_type`
  `required = TRUE/FALSE` setting as the source of truth. MCPR does not
  override manually declared schemas based on function defaults. This matches
  the schema-first pattern used by the official MCP TypeScript SDK.

For package authors using `inst/tools/` wrappers, the practical rule is simple:

```{r}
# Required in schema: no default in the wrapper
#' @param a numeric First operand (required).
#' @param b numeric Second operand (required).
add <- function(a, b) {
  mypackage::add(a = a, b = b)
}

# Optional in schema: defaulted in the wrapper
#' @param b numeric Multiplier (default: 1, i.e. identity).
multiply <- function(a, b = 1) {
  mypackage::multiply(a = a, b = b)
}
```

Two important caveats:

1. **Nullability is not the same as optionality.** A parameter like
   `contrast = NULL` is optional because the caller may omit it. Whether `NULL`
   is also a meaningful value is a separate concern handled by the tool
   implementation, not by the JSON Schema `required` array.
2. **Conditional requirements must still be documented in prose.** JSON Schema
   can express "always required" well, but many tools need rules like "required
   only when `what = \"help\"`" or "required for most plot types". Keep those
   constraints in the `@param` description and enforce them at runtime.

Place your tool files in `inst/tools/` (or any subdirectory convention you prefer) and point `ToolRegistry` at it. That's it — MCPR handles roxygen parsing, schema generation, JSON-RPC dispatch, private execution, and optional attachment when `manage_r_sessions` is registered.

**Note on paths:** `system.file(package = "pkg")` returns the *installed* package root, where `inst/` contents are lifted to the top level. During development with `devtools::load_all()`, tool files remain under `inst/` — use the source path directly in that case.

## Putting It All Together: A Package Entry Point

A downstream package typically exposes a convenience function:

```{r}
# In mypackage/R/server.R

#' Start the mypackage MCP Server
#' @param include_session_tools Include MCPR session management tools (default: FALSE)
#' @export
calc_server <- function(include_session_tools = FALSE) {
  dirs <- system.file("tools", package = "mypackage")

  if (include_session_tools) {
    dirs <- c(system.file(package = "MCPR"), dirs)
  }

  registry <- MCPR::ToolRegistry$new(tools_dir = dirs)
  registry$search_tools()

  if (include_session_tools) {
    # Keep session management, drop open-ended tools
    registry$filter(exclude = c("execute_r_code", "show_plot"))
  }

  # Ordinary package tools run in the private MCP server process.
  # If manage_r_sessions is included, attachment is controlled explicitly.
  MCPR::mcpr_server(registry = registry)
}
```

Users of the package get a focused, domain-specific MCP server — in this case exposing only the calculator tools (`add`, `subtract`, `multiply`, `divide`) plus optional attachment controls — without needing to understand MCPR internals.

## Summary

| I want to... | Code |
|---|---|
| Use MCPR defaults | `mcprServer$new()` |
| Add my tools alongside MCPR's | `ToolRegistry$new(tools_dir = c(mcpr, mine))` |
| Use only my tools | `ToolRegistry$new(tools_dir = mine)` |
| Cherry-pick from MCPR | `registry$filter(include = c("manage_r_sessions"))` |
| Remove dangerous tools | `registry$filter(exclude = c("execute_r_code"))` |
| Multiple tool subdirs | `tools_dir = c(dir1, dir2, dir3)` |
| See where tools came from | `registry$get_tool_summary()` or `tool$annotations$source_dir` |
